---
title: "Untitled"
author: "Hanna Berman"
date: "4/30/2021"
output: html_document
---

# Set up
## Packages
```{r}
library(tidyverse)
library(formattable)
theme_set(theme_bw())
```

## File paths

```{r}
unchGardSamples <- read_lines("/Users/hlberman/Desktop/unchGardSamples.txt")
```

### Abundance
```{r}
#Metaphlan Aabundance:
hmp2MetaphlanPath <- "../../humann2/merged_tables/hmp2mgs_mergedMetaphlanAbundance.tsv"
hmp2mDF <- hmp2MetaphlanPath %>%
  read_tsv() %>%
  filter(str_detect(ID, "s__"), !str_detect(ID, "t__")) %>%
  mutate(Species = str_extract(ID, "(?<=s__)\\w+$")) %>%
  select(Species, everything(), -ID) %>%
  as.data.frame() %>%
  dplyr::rename_at(2:ncol(.), ~str_extract(.x, "[0-9]{7}")) %>%
  column_to_rownames(var="Species") %>%
  t %>%
  as.data.frame() %>%
  mutate_all(funs(./100)) %>%
  rownames_to_column(var="SampleID")
hmp2mDF[1:6,1:5]

metaphlanGardAbundance <- hmp2mDF %>%
  select(SampleID, Gardnerella_vaginalis) %>%
  dplyr::rename(metaphlanGardAbundance=Gardnerella_vaginalis)

# Mapping Abundance
mapCladeAbundance <- "../hmp2gardCladeRelativeAbundance.tsv" %>%
  read_tsv()
head(mapCladeAbundance)
```

### Alignments
  Mapping
```{r}
mappingAlignments <- read_tsv("./hmp2cladeAlignmentsFinal.tsv")

mapCladeCounts <- mappingAlignments %>%
  group_by(SampleID) %>%
  count(Clade)
head(mapCladeCounts)

mapGardCounts <- mapCladeCounts %>%
  group_by(SampleID) %>%
  summarise(mapReads=sum(n)) %>%
  mutate(SampleID=as.character(SampleID))
head(mapGardCounts)  
```

  Metaphlan
```{r, message=FALSE}
metaphlanMarkerFiles <- list.files("/Users/hlberman/mnt/cluster/home6/hlberman/VMMG/humann2/hmp2mgs_gard_marker_counts", full.names = TRUE)
length(metaphlanMarkerFiles)

# metaphlanMarkerCounts <- metaphlanMarkerFiles %>%
#     map(read_tsv) %>%
#     map(., ~mutate(.x, Metaphlan2_Analysis=as.numeric(Metaphlan2_Analysis)))

metaphlanGardCounts <- "/Users/hlberman/mnt/cluster/home6/hlberman/VMMG/humann2/merged_tables/hmp2mgs_gardMarkerCounts.csv" %>%
  read_csv %>%
  mutate(SampleID=str_extract(SampleID, "(?<=SRR)[0-9]{7}"))

nrow(metaphlanGardCounts)
```

  Check for missing samples:
```{r}
missingMetaphlan <- setdiff(str_extract(metaphlanMarkerFiles, "[0-9]{7}(?=gard_marker_reads.tsv)"), metaphlanGardCounts$SampleID)
missingMetaphlan

metaphlanMarkerFiles %>%
  keep(~str_detect(.x, paste(missingMetaphlan, collapse = "|"))) %>%
  map(read_tsv)
```

  No *Gardnerlla* reads in sample `r missingMetaphlan`

  Set up tables for comparison
```{r}
compareDF <- data.frame(SampleID=missingMetaphlan, # add missing rows
                        metaphlanReads=0) %>%
  rbind(., metaphlanGardCounts) %>%
  full_join(mapGardCounts, by="SampleID") %>%
  replace_na(list(mapReads=0)) %>%
  full_join(metaphlanGardAbundance, by="SampleID")
head(compareDF)
```

# Reference genes
## Files
```{r}
refGard <- "../gard_phylogeny/GardnerellaMetadata.csv" %>%
  read_csv

refClades <- "../refGardnerellaCladesGenomos.csv" %>%
  read_csv
```
### Mapping References files
```{r}
refGenes <- "./ref_genes/refGenesDF.csv" %>%
  read_csv
```

### Metaphlan References
```{r}
# Genomes 
metaphlanMarkerGenomes <- "/Users/hlberman/mnt/cluster/home6/hlberman/Applications/metaphlan2/utils/species2genomes.txt" %>%
  read_lines() %>%
  .[str_detect(., "s__Gardnerella_vaginalis")] %>%
  str_split(., "\t") %>%
  unlist() %>%
  .[3:length(.)]
length(metaphlanMarkerGenomes)

# Genes
metaphlanMarkerIDs <-  "/Users/hlberman/mnt/cluster/home6/hlberman/Applications/metaphlan2/utils/gardnerella_vaginalis_marker_IDs.txt" %>%
  read_lines()
length(metaphlanMarkerIDs)
```
  36 genomes with 200 genes total

```{r}
metaphlanRefClades0 <- refGard %>%
  filter(str_extract(RefSeq_accession, "GCF_[0-9]{9}") %in% metaphlanMarkerGenomes) %>%
  left_join(refClades, by="Strain") %>%
  select(Strain, RefSeq_accession, in_phylogeny, Genomospecies, Clade)
head(metaphlanRefClades0)
```

  Strains that do not have a clade assignment:
```{r}
metaphlanRefClades0 %>%
  filter(in_phylogeny==FALSE)
```
  Know all bue 6420LIT from previous phylogeny. Look at Mash distances for 6420LIT placement
```{r}
#minhash distances from Mash
resultsPath <- "/Volumes/GoogleDrive/My Drive/Callahan Lab/metagenome_gardnerella/gardPhylogeny/assembly_quality_control/completeness_quality_dereplication"
load(file.path(resultsPath, "mashDists.Rdata"))
```

```{r}
GB6420Lit <- refGard %>% #genbank accession of 6420Lit
  filter(Strain=="6420LIT") %>%
  .$GenBank_accession
GB6420Lit

assembly1s <- mashDists %>%
  filter(assembly1==GB6420Lit) %>%
  select(-assembly1) %>%
  dplyr::rename(GenBank_accession=assembly2)

assembly2s <- mashDists %>%
  filter(assembly2==GB6420Lit) %>%
  select(-assembly2) %>%
  dplyr::rename(GenBank_accession=assembly1)

DF6420LitMash <- assembly1s %>%
  full_join(assembly2s, by = c("GenBank_accession", "distance", "p", "Matching_hashes")) %>%
  left_join(refGard[,c("Strain", "GenBank_accession")], by="GenBank_accession") %>%
  left_join(refClades, by="Strain") %>%
  filter(!is.na(Clade))

DF6420LitMash %>%
  filter(distance==min(distance))

DF6420LitMash %>%
  ggplot(aes(x=Clade, y=distance)) +
  geom_boxplot() +
  geom_jitter() +
  theme_bw()
```

6420Lit is from clade 4

```{r}
metaphlanRefClades <- metaphlanRefClades0 %>%
  mutate(Clade=case_when(Strain=="ATCC_14018"~"C1",
                         Strain=="JCP8481B"~"C3",
                         Strain=="JCP8017B"~"C2",
                         Strain=="6420LIT"~"C4",
                         Strain!="ATCC_14018|JCP8481B|JCP8017B|6420LIT"~Clade)) 
```

  Summarize clades represented in Metaphlan2 marker genes
```{r}
metaphlanRefClades %>%
  group_by(Clade) %>%
  tally %>%
  formattable()
```

  Summarize clades represented in mapping method marker genes
```{r}
refGenes %>%
  select(Strain, Clade) %>%
  unique() %>%
  group_by(Clade) %>%
  tally %>%
  formattable
```

# Compare alignments
```{r}
compareDF %>%
  gather("method", "reads", c(metaphlanReads, mapReads)) %>%
  ggplot(aes(x=reads, fill=method, color=method)) +
  geom_density(alpha=0.5) +
  scale_fill_discrete(labels=c("Our method reads", "Metaphlan2 reads")) +
  scale_color_discrete(labels=c("Our method reads", "Metaphlan2 reads"))

compareDF %>%
  gather("method", "reads", c(metaphlanReads, mapReads)) %>%
  ggplot(aes(x=reads, fill=method, color=method)) +
  geom_density(alpha=0.5) +
  xlim(0,5000) +
  scale_fill_discrete(labels=c("Our method reads", "Metaphlan2 reads")) +
  scale_color_discrete(labels=c("Our method reads", "Metaphlan2 reads"))

compareDF %>%
  ggplot(aes(x=mapReads, y=metaphlanReads)) +
  geom_abline(color="grey", linetype=2) +
  geom_point(alpha=0.5) +
  xlim(0,1000000) +
  coord_fixed() +
  labs(x="Aligned Marker Reads - Our Method", y="Aligned Marker Reads - Metaphlan2")
```

```{r}
compareDF %>%
  ggplot(aes(x=mapReads, y=metaphlanReads)) +
  geom_abline(color="grey", linetype=2) +
  geom_point(alpha=0.5) +
  labs(x="Aligned Marker Reads - Our Method", y="Aligned Marker Reads - Metaphlan2")
```
```{r}
lm(metaphlanReads~mapReads, data = compareDF)
```

```{r}
compareDF %>%
  mutate(unchGard=case_when(SampleID %in% unchGardSamples~"uncharacterized",
                            !(SampleID %in% unchGardSamples)&metaphlanGardAbundance!=0~"characterized",
                            metaphlanGardAbundance==0~"zero Gardnerella abundance")) %>%
  ggplot(aes(x=metaphlanReads, fill=unchGard)) +
  geom_histogram() +
  labs(x="Metaphlan Gardnerella Marker Reads", fill="Gardnerella clade resolution")
```

```{r}
compareDF %>%
  mutate(unchGard=case_when(SampleID %in% unchGardSamples~"uncharacterized",
                            !(SampleID %in% unchGardSamples)&metaphlanGardAbundance!=0~"characterized",
                            metaphlanGardAbundance==0~"zero Gardnerella abundance")) %>%
  ggplot(aes(x=metaphlanReads, fill=unchGard)) +
  geom_histogram() +
  xlim(0,1000) +
  labs(x="Metaphlan Gardnerella Marker Reads", fill="Gardnerella clade resolution")
```


  Uncharacterized *Gardnerella* samples only
```{r}
compareDF %>%
  filter(SampleID %in% unchGardSamples) %>%
  ggplot(aes(x=metaphlanReads)) +
  geom_histogram()
```


# Session info
```{r}
#save.image("/Users/hlberman/Desktop/saveImage.Rdata")
#load("/Users/hlberman/Desktop/saveImage.Rdata")
sessionInfo()
```