---
title: "Sample and Subject Metadata"
author: "HLB"
output:
  html_document: default
  html_notebook: default
---

The following script is to update metadata with detailed sample metadata and describe the sampling. 

# Setup
Packages
```{r}
library(tidyverse)
set.seed(7334)
```

# Collection Figure
```{r}
sgDF <- "./Tables_4_Ben.txt" %>%
  read_delim(delim="\t") %>%
  dplyr::rename(SampleIndex=`Sample Index`,
                SampleType=`Sample Type`,
                SubjectID=Subject,
                SampleID=`Sample ID`,
                GDcoll=GD_Coll,
                GWdell=GW_Dell,
                TermPre=`Term/Pre`,
                TotalReads=`Total reads`,
                TotalPairs=`Total pairs`,
                Sickle100_pe=`Sickle_100 (pe)`,
                Sickle_pairs=Sickle_pe,
                Human_pairs=Human_pe,
                PctTrim=`Pct Trim`,
                PctHuman=`Pct Human`) %>%
  mutate(SampleID=as.character(SampleID),
         SubjectID=as.character(SubjectID),
         PctTrim=as.numeric(as.character(str_extract(PctTrim, ".*(?=\\%)")))/100,
         PctHuman=as.numeric(as.character(str_extract(PctHuman, ".*(?=\\%)")))/100, 
         GWcoll=GDcoll/7,
         SampleID_a=paste0(substr(SampleID, 1, 5), "01", substr(SampleID, 8, 10)),
         RNAlater=case_when(str_detect(SampleID, "[0-9]{5}01[0-9]{3}")~"No",
                            str_detect(SampleID, "[0-9]{5}35[0-9]{3}")~"Yes"))
head(sgDF)
```

## sanity checks
```{r}
#sickle pairs == half of `Sickle100_pe`
sgDF %>%
  select(Sickle100_pe, Sickle_pairs) %>%
  mutate(test=Sickle_pairs*2,
         TF=case_when(test==Sickle100_pe~TRUE,
                      test!=Sickle100_pe~FALSE)) %>%
  .$TF %>%
  table()

#Percent trimmed is Sickle 100 paired and single reads divided by total reads
sgDF %>%
  select(TotalReads, Sickle100_pe, Sickle_sngl, PctTrim) %>%
  mutate(test=as.character(round(((Sickle100_pe+Sickle_sngl)/TotalReads), 4)),
         PctTrim=as.character(PctTrim),
         TF=case_when(test==PctTrim~TRUE,
                      test!=PctTrim~FALSE)) %>%
  .$TF %>%
  table()

#Percent human is bowtie paired and single reads divided by total reads
sgDF %>%
  select(Sickle100_pe, Sickle_pairs, Sickle_sngl, Human_pairs, Human_sngl,PctHuman) %>%
  mutate(test=as.character(round((Human_pairs*2+Human_sngl)/(Sickle100_pe+Sickle_sngl),4)),
         PctHuman=as.character(PctHuman),
         TF=case_when(test==PctHuman~TRUE,
                      test!=PctHuman~FALSE)) %>%
  .$TF %>%
  table
```
Unsure how percent human was calculated here...

```{r}
PS15 <- "./sampledata_VaginalSwabs_Pregnancy_DiGiulio2015.tsv" %>%
  read_tsv() %>%
  dplyr::rename(GDcoll=GDColl) %>%
  mutate(SubjectID= as.character(SubjectID),
         Cohort="Stanford",
         Method="Amplicon",
         PaperCohort="2015") %>%
  select(SampleID, SubjectID, GDcoll, Cohort, Method, PaperCohort)

PS17 <- "./MAP_MANUSCRIPT_IL02-IL09_ToBen_20170404_fixed.txt" %>%
  read_delim(delim="\t") %>%
  mutate(SubjectID= as.character(SubjectID),
         Cohort=case_when(UAB_Sample==0~"Stanford",
                                 UAB_Sample==1~"UAB"),
         Method="Amplicon",
         PaperCohort="2017") %>%
  select(SampleID, SubjectID, GDcoll, Cohort, Method, PaperCohort)

AMP <- rbind(PS15, PS17) %>%
  unique.data.frame()

sgDF0 <- sgDF %>%
  left_join(AMP[,c("SubjectID", "Cohort", "PaperCohort")], by="SubjectID") %>%
  filter(SampleID %in% sgDF$SampleID) %>%
  mutate(Cohort=replace_na(Cohort,"Stanford")) %>%
  unique.data.frame()


SHOT <- sgDF0 %>%
  select(SubjectID, GDcoll, Cohort, PaperCohort) %>%
  filter(PaperCohort=="2015"|PaperCohort=="2017") %>%
  mutate(Method="Shotgun") %>%
  unique.data.frame() 
```

# Collection Figure
```{r}
ggplot() +
  geom_point(data=AMP, aes(x=GDcoll, y=as.character(SubjectID), shape=Method, color=Cohort, size=Method), alpha=.5) +
   geom_point(data=SHOT, aes(x=GDcoll, y=as.character(SubjectID), shape=Method, size=Method), color="black") +
  xlim(0,280) +
     labs(x = "Gestation Days",
          y = "Subject") +
     scale_color_manual(values=c("olivedrab", "steelblue", "black"))+
     scale_shape_manual(values=c(19, 15)) +
     scale_size_manual(values=c(1,2)) +
     #theme_classic()+
     theme(axis.ticks.y = element_blank(),
     #axis.text.x = element_text(size = 10),
     axis.text.y  = element_blank()
     #axis.title.x = element_text(size=20),
     #axis.title.y = element_text(size=20)
     )
     #scale_x_continuous(breaks = c(0,50,100,150,200, 250))
#ggsave("./collectionPlot.svg", height=8, width=8.5)
```

# Trimesters
```{r}
sgDF1 <- sgDF0 %>%
  group_by(SubjectID) %>%
  mutate(sampleOrder=rank(GDcoll),
         sampleTrimester=case_when(GDcoll<=97~"First",
                                   GDcoll>=98&GDcoll<=195~"Second",
                                   GDcoll>=196~"Third")) %>%
  ungroup() 

#save updated dataframe
#write_tsv(sgDF1, "./shotgunMetadata.tsv")
```

# Samples/subjects per cohort
```{r}
sgDF1 %>%
  filter(Cohort=="UAB")%>%
  nrow()

sgDF1 %>%
  filter(Cohort=="UAB") %>%
  select(SubjectID) %>%
  n_distinct()

sgDF1 %>%
  filter(Cohort=="Stanford")%>%
  nrow()

sgDF1 %>%
  filter(Cohort=="Stanford") %>%
  select(SubjectID) %>%
  n_distinct()
```

# Subsetting
## Look at potential ways to subset samples
```{r}
range(sgDF1$GDcoll)
summary(sgDF1$GDcoll[sgDF1$sampleOrder==1])
summary(sgDF1$GDcoll[sgDF1$sampleOrder==2])
summary(sgDF1$GDcoll[sgDF1$sampleOrder==3])


sgDF1 %>%
  ggplot(aes(x=sampleOrder, y=GDcoll, group=sampleOrder)) +
  geom_boxplot() +
  geom_point() +
  facet_wrap(~Cohort) +
  labs(x="Sample Order", y="Gestation Days at Collection")

sgDF1 %>%
  ggplot(aes(x=GDcoll, y=as.character(SubjectID), color=Cohort)) +
  geom_point() +
  geom_vline(xintercept = 97.5)+
  geom_vline(xintercept = 195.5) +
  labs(x="Gestation Days at Collection", y="Subject ID")
```

## Randomly select one sample per subject from each subject
It seems that is the best way to maintain a small range of gestation days at collection of each range while excluding the least number of samples (will only exclude one sample)
```{r}
#Create dataframe of subsets
sgDFsubset <- sgDF1 %>%
  filter(sampleTrimester=="Second") %>%
  group_by(SubjectID) %>%
  sample_n(size=1) %>%
  ungroup

sampleSubset <- sgDFsubset$SampleID
sampleSubset
#save list of subset
#write_lines(sampleSubset, "./sampleSubset.txt")

#what subject is missing a sample
setdiff(unique(sgDF$SubjectID), sgDFsubset$SubjectID)
```

## gestation days range of subset
```{r}
summary(sgDFsubset$GDcoll)
```

# Session Info
```{r}
sessionInfo()
```