---
title: "Untitled"
author: "Hanna Berman"
date: "10/7/2019"
output: html_document
---

# Set up
## packages
```{r}
library(tidyverse)
library(phyloseq)
library(Biostrings)
```

```{r}
repo_dir <- "../../"

#shotgun metadata
sgDF <- file.path(repo_dir, "shotgunMetadata.tsv") %>%
  read_delim(delim = "\t") %>%
  mutate(SampleID=as.character(SampleID),
         SampleID_a=as.character(SampleID_a))
```

Set up metaphlan table
```{r}
load("./metaphlanAbundances.Rdata")

# #metaphlan data
# mDF0 <- file.path("/Volumes/GoogleDrive/My Drive/Callahan Lab/vagMicrobiomeMetagenome/humann2/merged_tables/mergedMetaphlanAbundance.tsv") %>%
#   read_tsv() %>%
#   filter(str_detect(ID, "s__"), !str_detect(ID, "t__")) %>%
#   mutate(Species = str_extract(ID, "(?<=s__)\\w+$")) %>%
#   select(Species, everything(), -ID) %>%
#   as.data.frame()
# 
# #make col and row names
# samps <- colnames(mDF0[2:ncol(mDF0)]) %>%
#   str_extract("[0-9]{10}")
# samps
# colnames(mDF0) <- c("Species", samps)
# rownames(mDF0) <- mDF0$Species
# 
# #Final edits to table
# mDF <- mDF0 %>%
#   select(-Species) %>%
#   t() %>%
#   as_tibble() %>%
#   mutate_all(funs(./100)) %>%
#   mutate(SampleID=samps) %>% 
#   select(SampleID, everything())
# mDF[1:6,1:5]
```

# Testing Absolute Abundance
## Number taxa vs filtered read count
```{r}
mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "TotalReads", "bbmapFiltered_pairs")], by="SampleID") %>%
  gather("measure", "readCount", c(TotalReads, bbmapFiltered_pairs)) %>%
  mutate(measure=case_when(measure=="TotalReads"~"Library Size",
                           measure=="bbmapFiltered_pairs"~"Bacterial Reads")) %>%
  ggplot(aes(x=readCount, y=nTaxa)) +
    geom_point() +
    geom_smooth(method="lm") +
    facet_wrap(~measure, scales="free_x") +
    labs(x="Reads", y="N Taxa")

mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "finalPctHuman_po", "finalPctHuman_all", "bbmapFiltered_pairs", "bbmapFiltered_sngl")], by="SampleID") %>%
  lm(data=., nTaxa~bbmapFiltered_pairs) %>%
  summary
```

```{r}
sgDF %>%
  ggplot(aes(x=TotalReads, y=(1-finalPctHuman_po))) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(x="Library Size", y="Percent Bacteria")
```

## Richness vs percent bacteria
```{r}
mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "finalPctHuman_po", "finalPctHuman_all", "bbmapFiltered_pairs", "bbmapFiltered_sngl")], by="SampleID") %>%
  ggplot(aes(x=(1-finalPctHuman_po)*100, y=nTaxa)) +
    geom_point() +
    geom_smooth(method="lm") +
    labs(x="Percent Bacteria", y="N Taxa")

mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "finalPctHuman_po", "finalPctHuman_all", "bbmapFiltered_pairs", "bbmapFiltered_sngl")], by="SampleID") %>%
  mutate(pct_bacteria=1-finalPctHuman_po) %>%
  lm(data=., nTaxa~pct_bacteria) %>%
  summary
```

# Make absolute abundance 
```{r}
# Total Gardnerella 
mDFaa <- mDF %>%
 left_join(sgDF[,c("SampleID", "finalPctHuman_po")], by="SampleID") %>%
 select(SampleID, finalPctHuman_po, everything()) %>%
 mutate_at(c(3:ncol(.)), ~.x*((1-finalPctHuman_po)/finalPctHuman_po)) #%>% #proportion 

# Gardnerella separated by clade
mDFaaG <- mDFG %>%
 left_join(sgDF[,c("SampleID", "finalPctHuman_po")], by="SampleID") %>%
 select(SampleID, finalPctHuman_po, everything()) %>%
 mutate_at(c(3:ncol(.)), ~.x*((1-finalPctHuman_po)/finalPctHuman_po)) #%>% #proportion 
```

# Figures
## Minimum Proportion vs. percent human
```{r}

```

## Correlation among important taxa
Organize data frame
```{r}
DF <- sgDF %>%
  select(SampleID, SubjectID, Cohort, GDcoll) %>%
  left_join(mDFaa[,c("SampleID", "Gardnerella_vaginalis")], by="SampleID") %>%
  left_join(mDFaaG[,c("SampleID",
                    "Gardnerella_vaginalis_C1", 
                    "Gardnerella_vaginalis_C2", 
                    "Gardnerella_vaginalis_C3", 
                    "Gardnerella_vaginalis_C4",
                    "Gardnerella_vaginalis_C5",
                    "Gardnerella_vaginalis_C6",
                    "Lactobacillus_crispatus", 
                    "Lactobacillus_gasseri",
                    "Lactobacillus_jensenii", 
                    "Lactobacillus_iners", 
                    "Atopobium_vaginae", 
                    "Mycoplasma_hominis",
                    "Prevotella_bivia",
                    "Ureaplasma_parvum")],
            by="SampleID") %>%
    dplyr::rename(C1=Gardnerella_vaginalis_C1,
                C2=Gardnerella_vaginalis_C2,
                C3=Gardnerella_vaginalis_C3,
                C4=Gardnerella_vaginalis_C4,
                C5=Gardnerella_vaginalis_C5,
                C6=Gardnerella_vaginalis_C6,
                G_vaginalis=Gardnerella_vaginalis,
                A_vaginae=Atopobium_vaginae,
                L_crispatus=Lactobacillus_crispatus,
                L_gasseri=Lactobacillus_gasseri,
                L_iners=Lactobacillus_iners,
                L_jensenii=Lactobacillus_jensenii,
                P_bivia=Prevotella_bivia,
                U_parvum=Ureaplasma_parvum,
                M_hominis=Mycoplasma_hominis) %>%
  mutate(SampleID=as.character(SampleID),
         SubjectID=as.character(SubjectID),
         GDcoll=as.character(GDcoll))

subsetSamples <- read_lines(file.path("..", "..", "sampleSubset.txt"))
```

Plot
```{r, fig.height=8, fig.width=8}
DF0 <- DF %>%
  mutate(xC1=C1, 
         xC2=C2,
         xC3=C3,
         xC4=C4,
         xC5=C5,
         xC6=C6,
         xG_vaginalis=G_vaginalis,
         xA_vaginae=A_vaginae,
         xL_crispatus=L_crispatus,
         xL_gasseri=L_gasseri, 
         xL_iners=L_iners, 
         xL_jensenii=L_jensenii, 
         xP_bivia=P_bivia, 
         xU_parvum=U_parvum, 
         xM_hominis=M_hominis) %>%
  gather("Taxon1", "absoluteAbundanceTax1", c(C1, C2, C3, C4, C5, C6, G_vaginalis, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
  gather("Taxon2", "absoluteAbundanceTax2", c(xC1, xC2, xC3, xC4, xC5, xC6, xG_vaginalis, xA_vaginae, xL_crispatus, xL_gasseri, xL_iners, xL_jensenii, xP_bivia, xU_parvum, xM_hominis)) %>%
  mutate(Taxon2=case_when(Taxon2=="xC1"~"C1", 
                          Taxon2=="xC2"~"C2",
                          Taxon2=="xC3"~"C3",
                          Taxon2=="xC4"~"C4",
                          Taxon2=="xC5"~"C5",
                          Taxon2=="xC6"~"C6",
                          Taxon2=="xG_vaginalis"~"G_vaginalis",
                          Taxon2=="xA_vaginae"~"A_vaginae",
                          Taxon2=="xL_crispatus"~"L_crispatus",
                          Taxon2=="xL_gasseri"~"L_gasseri", 
                          Taxon2=="xL_iners"~"L_iners", 
                          Taxon2=="xL_jensenii"~"L_jensenii", 
                          Taxon2=="xP_bivia"~"P_bivia", 
                          Taxon2=="xU_parvum"~"U_parvum", 
                          Taxon2=="xM_homonis"~"M_hominis")) #%>%
  #filter(Taxon1!=Taxon2)
  
DF0 %>%
  filter(Taxon1!="G_vaginalis",
         Taxon2!="G_vaginalis",
         Cohort=="Stanford") %>%
ggplot(aes(x=absoluteAbundanceTax1, y=absoluteAbundanceTax2)) +
  #geom_abline(intercept = 0, slope = 1, color="grey", linetype=2) +
  geom_smooth(method="lm", se=FALSE, linetype=1, color="blue") +
  geom_point() +
  facet_grid(Taxon1~Taxon2) +
  coord_fixed() +
  scale_y_log10() +
  scale_x_log10() +
  theme(axis.text.x = element_text(angle=90)) 
  #xlim(0,1) +
  #ylim(0,1)
```

Correlation Coefficients
```{r}
# UAB
DF %>%
  filter(Cohort=="UAB") %>%
  select(-SampleID, -SubjectID, -GDcoll, -Cohort) %>%
  cor %>%
  View

# Stanford
DF %>%
  filter(Cohort=="Stanford") %>%
  select(-SampleID, -SubjectID, -GDcoll, -Cohort) %>%
  cor %>%
  View
```

Clade presence Absence by Percent Bacterial Reads
```{r}
DF %>%
  gather("Taxon", "absoluteAbundance", c(C1, C2, C3, C4, C5, C6, G_vaginalis, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
  mutate(absoluteAbundance=case_when(absoluteAbundance==0~"absent",
                                     absoluteAbundance>0~"present")) %>%
  left_join(sgDF[,c("SampleID", "Sickle_pairs", "bbmapFiltered_pairs", "finalPctHuman_po")]) %>%
  ggplot(aes(x=absoluteAbundance, y=1-finalPctHuman_po)) +
  geom_boxplot() +
  geom_jitter() +
  facet_grid(Cohort~Taxon) +
  theme(axis.text.x = element_text(angle=90))

cladesByPctBact <- DF %>%
  mutate_at(vars(G_vaginalis, C1, C2, C3, C4, C5, C6), ~case_when(.x==0~"absent",
                                                     .x>0~"present")) %>%
  left_join(sgDF[,c("SampleID", "Sickle_pairs", "bbmapFiltered_pairs", "finalPctHuman_po")]) %>%
  mutate(ratio=((1-finalPctHuman_po)/finalPctHuman_po),
         pctBact=1-finalPctHuman_po) %>%
  filter(Cohort=="Stanford")

tTestByPctBact <- map2(list(cladesByPctBact$pctBact, cladesByPctBact$pctBact, cladesByPctBact$pctBact, cladesByPctBact$pctBact, cladesByPctBact$pctBact, cladesByPctBact$pctBact, cladesByPctBact$pctBact), list(cladesByPctBact$G_vaginalis, cladesByPctBact$C1, cladesByPctBact$C2, cladesByPctBact$C3, cladesByPctBact$C4, cladesByPctBact$C5, cladesByPctBact$C6), ~t.test(.x~.y, alternative="less"))
tTestByPctBact
```
Significance at a=0.05
```{r}
map(tTestByPctBact, 3) %>%
  map(., ~.x<0.05)

map(tTestByPctBact, 3) %>%
  map(., ~.x<(0.05/7))
```
Clade presence absence by total abundance
```{r}
mDFaa %>%
  select(-finalPctHuman_po) %>%
  gather("tax", "ab", matches("._.")) %>%
  group_by(SampleID) %>%
  summarise(total_abundance=sum(ab)) %>%
  select(SampleID, total_abundance) %>%
  left_join(DF, by="SampleID") %>%
  gather("Taxon", "absoluteAbundance", c(C1, C2, C3, C4, C5, C6, G_vaginalis, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
  mutate(absoluteAbundance=case_when(absoluteAbundance==0~"absent",
                                     absoluteAbundance>0~"present")) %>%
  ggplot(aes(x=absoluteAbundance, y=total_abundance)) +
  geom_boxplot() +
  geom_jitter() +
  facet_grid(Cohort~Taxon) +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle=90))

cladesByTotAb <- mDFaa %>%
  select(-finalPctHuman_po) %>%
  gather("tax", "ab", matches("._.")) %>%
  group_by(SampleID) %>%
  summarise(total_abundance=sum(ab)) %>%
  select(SampleID, total_abundance) %>%
  left_join(DF, by="SampleID") %>%
  mutate_at(vars(G_vaginalis, C1, C2, C3, C4, C5, C6), ~case_when(.x==0~"absent",
                                                     .x>0~"present")) %>%
  filter(Cohort=="UAB")

tTestByTotAb <- map2(list(cladesByTotAb$total_abundance, cladesByTotAb$total_abundance, cladesByTotAb$total_abundance, cladesByTotAb$total_abundance, cladesByTotAb$total_abundance, cladesByTotAb$total_abundance, cladesByTotAb$total_abundance), list(cladesByTotAb$G_vaginalis, cladesByTotAb$C1, cladesByTotAb$C2, cladesByTotAb$C3, cladesByTotAb$C4, cladesByTotAb$C5, cladesByTotAb$C6), ~t.test(.x~.y, alternative="less"))
tTestByTotAb
```
Significance by a=0.05
```{r}
map(tTestByTotAb, 3) %>%
  map(., ~.x<0.05)

map(tTestByTotAb, 3) %>%
  map(., ~.x<(0.05/7))
```

# Session Info
```{r}
sessionInfo()
```