---
title: "Absolute Abundance"
author: "Hanna Berman"
date: "10/7/2019"
output: html_document
---

# Set up
## packages
```{r}
library(tidyverse)
library(phyloseq)
library(Biostrings)
library(broom)
`%!in%` <- Negate(`%in%`)
```

## Paths and values
```{r}
repo_dir <- "../../"

#shotgun metadata
sgDF <- file.path(repo_dir, "shotgunMetadata.tsv") %>%
  read_delim(delim = "\t") %>%
  mutate(SampleID=as.character(SampleID),
         SampleID_a=as.character(SampleID_a))

taxaOfInterest <- c("Gardnerella_vaginalis_C1", 
                    "Gardnerella_vaginalis_C2", 
                    "Gardnerella_vaginalis_C3", 
                    "Gardnerella_vaginalis_C4",
                    "Gardnerella_vaginalis_C5",
                    "Gardnerella_vaginalis_C6",
                    "Gardnerella_vaginalis_Uncharacterized",
                    "Lactobacillus_crispatus", 
                    "Lactobacillus_gasseri",
                    "Lactobacillus_jensenii", 
                    "Lactobacillus_iners", 
                    "Atopobium_vaginae", 
                    "Mycoplasma_hominis",
                    "Prevotella_bivia",
                    "Ureaplasma_parvum")
taxaOfInterestAbbr <- c("C1", 
                        "C2", 
                        "C3", 
                        "C4",
                        "C5",
                        "C6",
                        "Gv_uncharacterized",
                        "G_vaginalis",
                        "L_crispatus", 
                        "L_gasseri",
                        "L_jensenii", 
                        "L_iners", 
                        "A_vaginae", 
                        "M_hominis",
                        "P_bivia",
                        "U_parvum")

Gvaginalis <- c("C1", "C2", "C3", "C4","C5", "C6", "Gv_uncharacterized", "G_vaginalis")

 cladeColors <- list(scale_color_manual(values=c("#CC79A7", "#D55E00", "#56B4E9", "#0072B2", "#E69F00", "#009E73", "#A9A9A9", "#000000"), labels=c("C1", "C2", "C3", "C4", "C5", "C6", "Uncharacterized", bquote('Total'~italic(Gardnerella)))),
                     scale_fill_manual(values=c("#CC79A7", "#D55E00", "#56B4E9", "#0072B2", "#E69F00", "#009E73", "#A9A9A9", "#000000"), labels=c("C1", "C2", "C3", "C4", "C5", "C6", "Uncharacterized", bquote('Total'~italic(Gardnerella)))))
```

## Set up metaphlan table
```{r}
load("./metaphlanAbundances.Rdata")

# #metaphlan data
# mDF0 <- file.path("/Volumes/GoogleDrive/My Drive/Callahan Lab/vagMicrobiomeMetagenome/humann2/merged_tables/mergedMetaphlanAbundance.tsv") %>%
#   read_tsv() %>%
#   filter(str_detect(ID, "s__"), !str_detect(ID, "t__")) %>%
#   mutate(Species = str_extract(ID, "(?<=s__)\\w+$")) %>%
#   select(Species, everything(), -ID) %>%
#   as.data.frame()
# 
# #make col and row names
# samps <- colnames(mDF0[2:ncol(mDF0)]) %>%
#   str_extract("[0-9]{10}")
# samps
# colnames(mDF0) <- c("Species", samps)
# rownames(mDF0) <- mDF0$Species
# 
# #Final edits to table
# mDF <- mDF0 %>%
#   select(-Species) %>%
#   t() %>%
#   as_tibble() %>%
#   mutate_all(funs(./100)) %>%
#   mutate(SampleID=samps) %>% 
#   select(SampleID, everything())
# mDF[1:6,1:5]
```

# Testing Absolute Abundance
## Number taxa vs filtered read count
```{r}
mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "TotalReads", "bbmapFiltered_pairs")], by="SampleID") %>%
  gather("measure", "readCount", c(TotalReads, bbmapFiltered_pairs)) %>%
  mutate(measure=case_when(measure=="TotalReads"~"Library Size",
                           measure=="bbmapFiltered_pairs"~"Bacterial Reads")) %>%
  ggplot(aes(x=readCount, y=nTaxa)) +
    geom_point() +
    geom_smooth(method="lm") +
    facet_wrap(~measure, scales="free_x") +
    labs(x="Reads", y="N Taxa")

mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "finalPctHuman_po", "finalPctHuman_all", "bbmapFiltered_pairs", "bbmapFiltered_sngl")], by="SampleID") %>%
  lm(data=., nTaxa~bbmapFiltered_pairs) %>%
  summary
```

```{r}
sgDF %>%
  ggplot(aes(x=TotalReads, y=(1-finalPctHuman_po))) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(x="Library Size", y="Percent Bacteria")
```

## Richness vs percent bacteria
```{r}
mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "finalPctHuman_po", "finalPctHuman_all", "bbmapFiltered_pairs", "bbmapFiltered_sngl")], by="SampleID") %>%
  ggplot(aes(x=(1-finalPctHuman_po)*100, y=nTaxa)) +
    geom_point() +
    geom_smooth(method="lm") +
    labs(x="Percent Bacteria", y="N Taxa")

mDF %>%
  gather("Taxon", "Abundance", c(2:ncol(.))) %>%
  filter(Abundance>0) %>%
  group_by(SampleID) %>%
  summarise(nTaxa=n_distinct(Taxon)) %>%
  left_join(sgDF[,c("SampleID", "finalPctHuman_po", "finalPctHuman_all", "bbmapFiltered_pairs", "bbmapFiltered_sngl")], by="SampleID") %>%
  mutate(pct_bacteria=1-finalPctHuman_po) %>%
  lm(data=., nTaxa~pct_bacteria) %>%
  summary
```

# Make absolute abundance 
```{r}
# Total Gardnerella 
mDFaa <- mDF %>%
 left_join(sgDF[,c("SampleID", "finalPctHuman_po")], by="SampleID") %>%
 select(SampleID, finalPctHuman_po, everything()) %>%
 mutate_at(c(3:ncol(.)), ~.x*((1-finalPctHuman_po)/finalPctHuman_po)) #%>% #proportion 

# Gardnerella separated by clade
mDFaaG <- mDFG %>%
 left_join(sgDF[,c("SampleID", "finalPctHuman_po")], by="SampleID") %>%
 select(SampleID, finalPctHuman_po, everything()) %>%
 mutate_at(c(3:ncol(.)), ~.x*((1-finalPctHuman_po)/finalPctHuman_po)) #%>% #proportion 
```


# Analyses
Organize data frame
```{r}
# dataframe with each taxon as a column
DF <- sgDF %>%
  select(SampleID, SubjectID, Cohort, GDcoll) %>%
  left_join(mDFaa[,c("SampleID", "Gardnerella_vaginalis")], by="SampleID") %>%
  left_join(mDFaaG[,c("SampleID", taxaOfInterest)], by="SampleID") %>%
  dplyr::rename(C1=Gardnerella_vaginalis_C1,
                C2=Gardnerella_vaginalis_C2,
                C3=Gardnerella_vaginalis_C3,
                C4=Gardnerella_vaginalis_C4,
                C5=Gardnerella_vaginalis_C5,
                C6=Gardnerella_vaginalis_C6,
                Gv_uncharacterized=Gardnerella_vaginalis_Uncharacterized,
                G_vaginalis=Gardnerella_vaginalis,
                A_vaginae=Atopobium_vaginae,
                L_crispatus=Lactobacillus_crispatus,
                L_gasseri=Lactobacillus_gasseri,
                L_iners=Lactobacillus_iners,
                L_jensenii=Lactobacillus_jensenii,
                P_bivia=Prevotella_bivia,
                U_parvum=Ureaplasma_parvum,
                M_hominis=Mycoplasma_hominis) %>%
  mutate(SampleID=as.character(SampleID),
         SubjectID=as.character(SubjectID),
         GDcoll=as.character(GDcoll))

totalAbundances <- sgDF %>%
  mutate(totalAbundance=(1-finalPctHuman_po)/finalPctHuman_po) %>%
  select(SampleID, totalAbundance) 


subsetSamples <- read_lines(file.path("..", "..", "sampleSubset.txt"))
```


## Distribution of each clade and total *G. vaginalis*
```{r}
aaCladesPlot <- DF %>%
  left_join(totalAbundances, by="SampleID") %>%
  gather("Clade", "Abundance", c(G_vaginalis, C1, C2, C3, C4, C5, C6, Gv_uncharacterized)) %>%
  mutate(Clade=factor(Clade, levels=Gvaginalis)) %>%
  ggplot(aes(x=Clade, y=Abundance, color=Clade)) +
    geom_boxplot(outlier.alpha=0) +
    geom_jitter() +
    cladeColors +
    #scale_color_manual(values=c("#CC79A7", "#D55E00", "#56B4E9", "#0072B2", "#E69F00", "#009E73", "#000000"), labels=c("C1", "C2", "C3", "C4", "C5", "C6", "Uncharacterized", bquote('Total'~italic(Gardnerella)))) +
    facet_grid(Cohort~.)

aaCladesPlot
aaCladesPlot +
  ylim(0,1)
aaCladesPlot +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Correlation among important taxa
### Scatterplots
```{r, fig.height=8, fig.width=8, warning=FALSE}
#Taxon 1 vs. Taxon 2
DF0 <- DF %>%
  mutate(xC1=C1, 
         xC2=C2,
         xC3=C3,
         xC4=C4,
         xC5=C5,
         xC6=C6,
         xGv_uncharacterized=Gv_uncharacterized,
         xG_vaginalis=G_vaginalis,
         xA_vaginae=A_vaginae,
         xL_crispatus=L_crispatus,
         xL_gasseri=L_gasseri, 
         xL_iners=L_iners, 
         xL_jensenii=L_jensenii, 
         xP_bivia=P_bivia, 
         xU_parvum=U_parvum, 
         xM_hominis=M_hominis) %>%
  gather("Taxon1", "absoluteAbundanceTax1", c(C1, C2, C3, C4, C5, C6, Gv_uncharacterized, G_vaginalis, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
  gather("Taxon2", "absoluteAbundanceTax2", c(xC1, xC2, xC3, xC4, xC5, xC6, xGv_uncharacterized, xG_vaginalis, xA_vaginae, xL_crispatus, xL_gasseri, xL_iners, xL_jensenii, xP_bivia, xU_parvum, xM_hominis)) %>%
  mutate(Taxon2=case_when(Taxon2=="xC1"~"C1", 
                          Taxon2=="xC2"~"C2",
                          Taxon2=="xC3"~"C3",
                          Taxon2=="xC4"~"C4",
                          Taxon2=="xC5"~"C5",
                          Taxon2=="xC6"~"C6",
                          Taxon2=="xGv_uncharacterized"~"Gv_uncharacterized",
                          Taxon2=="xG_vaginalis"~"G_vaginalis",
                          Taxon2=="xA_vaginae"~"A_vaginae",
                          Taxon2=="xL_crispatus"~"L_crispatus",
                          Taxon2=="xL_gasseri"~"L_gasseri", 
                          Taxon2=="xL_iners"~"L_iners", 
                          Taxon2=="xL_jensenii"~"L_jensenii", 
                          Taxon2=="xP_bivia"~"P_bivia", 
                          Taxon2=="xU_parvum"~"U_parvum", 
                          Taxon2=="xM_hominis"~"M_hominis"))
  
DF0 %>%
  split(.$Cohort) %>%
  #map(~filter_at(.x, vars(Taxon1, Taxon2), ~.x!="G_vaginalis")) %>% # remove G. vaginalis from list
  map(~ggplot(.x, aes(x=absoluteAbundanceTax1, y=absoluteAbundanceTax2)) +
    #geom_abline(intercept = 0, slope = 1, color="grey", linetype=2) +
    geom_smooth(method="lm", se=FALSE, linetype=1, color="blue") +
    geom_point() +
    facet_grid(Taxon1~Taxon2) +
    coord_fixed() +
    scale_y_log10() +
    scale_x_log10() +
    theme(axis.text.x = element_text(angle=90)) +
    labs(title=paste(unique(.x$Cohort), "Cohort"), x="Absolute abundance: Taxon 1", y="Absolute abundance: Taxon 2"))
```

### Correlations
```{r, message=FALSE, warning=FALSE, fig.height=3, fig.width=3}
#function to calcualte spearman's rank correlation coefficients
corTestFig <- function(df, test, name){
  cor_test_results <- df %>%
    split(list(.$Taxon1, .$Taxon2, .$Cohort)) %>%
      map(~cor.test(~absoluteAbundanceTax1+absoluteAbundanceTax2, data=.x, alternative="two.sided", method=test)) %>%
      map(tidy) %>%
      map2(., names(.), ~mutate(.x, Group=.y)) %>%
      purrr::reduce(full_join, by = c("estimate", "statistic", "p.value", "method", "alternative", "Group")) %>%
      separate(Group, c("Taxon1", "Taxon2", "Cohort"), sep="\\.") %>%
      select(-p.value)
  plot <- cor_test_results %>%
## Remove diagonal or GvxGv
    filter(Taxon1!=Taxon2,
           Taxon1 == "G_vaginalis" & Taxon2 %!in% Gvaginalis|
           Taxon1 == "Gv_uncharacterized" & Taxon2 %!in% Gvaginalis|
           Taxon1 %in% c("C1", "C2", "C3", "C4", "C5", "C6", "L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum"),   
           Taxon2 == "G_vaginalis" & Taxon1 %!in% Gvaginalis|
           Taxon2 == "Gv_uncharacterized" & Taxon1 %!in% Gvaginalis|
           Taxon2 %in% c("C1", "C2", "C3", "C4", "C5", "C6", "L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")) %>%
## Remove other side of diagonal       
    # filter(Taxon1!=Taxon2,
    #         Taxon1 == "G_vaginalis" & Taxon2 %in% c("L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum") |
    #         Taxon1 == "C1" & Taxon2 %in% taxaOfInterestAbbr |
    #         Taxon1 == "C2" & Taxon2 %in% c("C3","C4","C5","C6","L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum") |
    #         Taxon1 == "C3" & Taxon2 %in% c("C4","C5","C6","L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum") |
    #         Taxon1 == "C4" & Taxon2 %in% c("C5","C6","L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "C5" & Taxon2 %in% c("C6",                                                                       "L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "C6" & Taxon2 %in% c("L_crispatus", "L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "L_crispatus" & Taxon2 %in% c("L_gasseri","L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "L_gasseri" & Taxon2 %in% c("L_jensenii", "L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "L_jensenii" & Taxon2 %in% c("L_iners", "A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "L_iners" & Taxon2 %in% c("A_vaginae", "M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "A_vaginae" & Taxon2 %in% c("M_hominis", "P_bivia", "U_parvum")|
    #         Taxon1 == "M_hominis" & Taxon2 %in% c("P_bivia", "U_parvum")|
    #         Taxon1 == "P_bivia" & Taxon2 == "U_parvum") %>%
      mutate(
        # p.value=case_when(p.value>0.01~"",
        #                      p.value<=0.01&p.value>0.001~"*",
        #                      p.value<=0.001&p.value>0.0008~"**",
        #                      p.value<=0.0008&p.value>0.0001~"***",
        #                      p.value<=0.0001~"****"),
          Taxon1=factor(Taxon1, levels=taxaOfInterestAbbr),
          Taxon2=factor(Taxon2, levels = rev(taxaOfInterestAbbr))) %>%
    #filter(Cohort=="Stanford") %>%
    ggplot(aes(x=Taxon1, y=Taxon2, fill=estimate)) + #label=p.value
      geom_tile() +
      #geom_text() +
      scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
      scale_x_discrete(drop=FALSE) +
      scale_y_discrete(drop=FALSE) +
      facet_wrap(~Cohort) +
      theme(axis.text.x = element_text(angle=90)) +
      labs(x="Taxon 1 ", y="Taxon 2", title = paste("Abundance Correlations:", test), subtitle = name)
  outputList <- list(Results=cor_test_results, Figure=plot)   
    return(outputList)
}

#list(AllSamples=DF0, SampleSubset=filter(DF0, SampleID %in% subsetSamples)) %>% # full data frame and data frame with only subset samples
#   map2(c("All Samples", "Sample Subset"), corTestFig) # run function
```

#### Absolute Abundance
````{r, warning=FALSE}
pmap(list(list(AllSamples=DF0, SampleSubset=filter(DF0, SampleID %in% subsetSamples), AllSamples=DF0, SampleSubset=filter(DF0, SampleID %in% subsetSamples)),
          list("spearman","spearman", "pearson", "pearson"),
          list("Absolute Abundance: All Samples", "Absolute Abundance: Sample Subset", "Absolute Abundance: All Samples", "Absolute Abundance: Sample Subset")),
     corTestFig)
```


#### Relative Abundance
```{r, warning=FALSE}
#make data frame
DFrr0 <- sgDF %>%
  select(SampleID, SubjectID, Cohort, GDcoll) %>%
  left_join(mDF[,c("SampleID", "Gardnerella_vaginalis")], by="SampleID") %>%
  left_join(mDFG[,c("SampleID", taxaOfInterest)], by="SampleID") %>%
  dplyr::rename(C1=Gardnerella_vaginalis_C1,
                C2=Gardnerella_vaginalis_C2,
                C3=Gardnerella_vaginalis_C3,
                C4=Gardnerella_vaginalis_C4,
                C5=Gardnerella_vaginalis_C5,
                C6=Gardnerella_vaginalis_C6,
                Gv_uncharacterized=Gardnerella_vaginalis_Uncharacterized,
                G_vaginalis=Gardnerella_vaginalis,
                A_vaginae=Atopobium_vaginae,
                L_crispatus=Lactobacillus_crispatus,
                L_gasseri=Lactobacillus_gasseri,
                L_iners=Lactobacillus_iners,
                L_jensenii=Lactobacillus_jensenii,
                P_bivia=Prevotella_bivia,
                U_parvum=Ureaplasma_parvum,
                M_hominis=Mycoplasma_hominis) %>%
  mutate(SampleID=as.character(SampleID),
         SubjectID=as.character(SubjectID),
         GDcoll=as.character(GDcoll))

DFrr <- DFrr0 %>%
  mutate(xC1=C1, 
         xC2=C2,
         xC3=C3,
         xC4=C4,
         xC5=C5,
         xC6=C6,
         xGv_uncharacterized=Gv_uncharacterized,
         xG_vaginalis=G_vaginalis,
         xA_vaginae=A_vaginae,
         xL_crispatus=L_crispatus,
         xL_gasseri=L_gasseri, 
         xL_iners=L_iners, 
         xL_jensenii=L_jensenii, 
         xP_bivia=P_bivia, 
         xU_parvum=U_parvum, 
         xM_hominis=M_hominis) %>%
  gather("Taxon1", "absoluteAbundanceTax1", c(C1, C2, C3, C4, C5, C6, Gv_uncharacterized, G_vaginalis, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
  gather("Taxon2", "absoluteAbundanceTax2", c(xC1, xC2, xC3, xC4, xC5, xC6, xGv_uncharacterized, xG_vaginalis, xA_vaginae, xL_crispatus, xL_gasseri, xL_iners, xL_jensenii, xP_bivia, xU_parvum, xM_hominis)) %>%
  mutate(Taxon2=case_when(Taxon2=="xC1"~"C1", 
                          Taxon2=="xC2"~"C2",
                          Taxon2=="xC3"~"C3",
                          Taxon2=="xC4"~"C4",
                          Taxon2=="xC5"~"C5",
                          Taxon2=="xC6"~"C6",
                          Taxon2=="xGv_uncharacterized"~"Gv_uncharacterized",
                          Taxon2=="xG_vaginalis"~"G_vaginalis",
                          Taxon2=="xA_vaginae"~"A_vaginae",
                          Taxon2=="xL_crispatus"~"L_crispatus",
                          Taxon2=="xL_gasseri"~"L_gasseri", 
                          Taxon2=="xL_iners"~"L_iners", 
                          Taxon2=="xL_jensenii"~"L_jensenii", 
                          Taxon2=="xP_bivia"~"P_bivia", 
                          Taxon2=="xU_parvum"~"U_parvum", 
                          Taxon2=="xM_hominis"~"M_hominis"))
#Run tests  
pmap(list(list(AllSamples=DF0, SampleSubset=filter(DFrr, SampleID %in% subsetSamples), AllSamples=DFrr, SampleSubset=filter(DFrr, SampleID %in% subsetSamples)),
          list("spearman","spearman", "pearson", "pearson"),
          list("Relatve Abundance: All Samples", "Relative Abundance: Sample Subset", "Relative Abundance: All Samples", "Relative Abundance: Sample Subset")),
     corTestFig)
```

## Clade presence-absence vs. total abundance
```{r, message=FALSE}
# # safeWilcox <- safely(wilcox.test)
# PresAbsVsAbundance <- function(df, name){
#  df0 <- df %>%
#     left_join(totalAbundances, by="SampleID") %>%
#     gather("Taxon", "presAbs", c(G_vaginalis, Gv_uncharacterized, C1, C2, C3, C4, C5, C6, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
#     mutate(presAbs=case_when(presAbs>0~"present",
#                              presAbs==0~"absent"),
#            presAbs=factor(presAbs, levels=c("present", "absent")),
#            Taxon=factor(Taxon, levels=taxaOfInterestAbbr, labels=c("C1", "C2", "C3", "C4", "C5", "C6", "GvU", "Gv", "Lc", "Lg", "Lj", "Li", "Av", "Mh", "Pb", "Up")))
#   # wilcoxResults <- df0 %>%
#   #   split(list(.$Taxon, .$Cohort)) %>% #split
#   #   map(~safeWilcox(totalAbundance~presAbs, data=.x, alternative="two.sided")) %>% # run wilcox test
#   #   map("result") %>%
#   #   map(~tidy(.x)) %>%
#   #   map2(., names(.), ~mutate(.x, Taxon=.y)) %>%  # add name to table
#   #   purrr::reduce(full_join) %>%
#   #   separate(Taxon, c("Taxon", "Cohort"), sep = "\\.") %>%
#   #   mutate(BH=p.adjust(p.value, method = "BH")) %>%
#   #   select(Cohort, Taxon, method, alternative, statistic, p.value, BH) # re-order columns for easier reading
#   
#   plot <- df0 %>%
#     ggplot(aes(x=presAbs, y=totalAbundance)) +
#       geom_boxplot(outlier.alpha = 0) +
#       geom_jitter() +
#       facet_grid(Cohort~Taxon) +
#       scale_x_discrete(drop=FALSE) +
#       scale_y_log10() +
#       theme(axis.text.x = element_text(angle=90)) +
#       labs(x=element_blank(), y="Bacterial Load", title="Presence-Absence vs. Abundance", subtitle = name)
#   outputList <- list(Figure=plot) #`Wilcox Results`=wilcoxResults, 
# }
# 
# list(AllSamples=DF, SampleSubset=filter(DF, SampleID %in% subsetSamples)) %>% # full data frame and data frame with only subset samples
#    map2(c("All Samples", "Sample Subset"), PresAbsVsAbundance) # run function
```

# Prelim/F31 figures
## Heat map
All Samples
```{r, message=FALSE, warning=FALSE, echo=FALSE}
# corTestFig1(DF1)
```

Sample Subset
```{r, message=FALSE, warning=FALSE, echo=FALSE}
# corTestFig1(DF2)
```

## bacterial load vs. presence absence
```{r, echo=FALSE}
# prelimPresAbsAll <- DF %>%
#     left_join(totalAbundances, by="SampleID") %>%
#     gather("Taxon", "presAbs", c(G_vaginalis, C1, C2, C3, C4, C5, C6, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
#     filter(Taxon %in% c("G_vaginalis", "C1", "C2", "C3", "C4", "C5", "C6", "L_crispatus", "L_gasseri", "L_jensenii", "L_iners")) %>%
#     mutate(Taxon = factor(Taxon, levels=c("G_vaginalis", "C1", "C2", "C3", "C4", "C5", "C6", "L_crispatus", "L_gasseri", "L_jensenii", "L_iners"), labels=taxaOfInterestAbbr0),
#            presAbs=case_when(presAbs>0~"present", 
#                              presAbs==0~"absent"),
#            presAbs=factor(presAbs, levels=c("present", "absent")))
#   
#  prelimPresAbsSubset <-  prelimPresAbsAll %>%
#   filter(SampleID %in% subsetSamples)
# 
#   prelimPresAbsFigure <- function(df){
#       ggplot(data=df, aes(x=presAbs, y=totalAbundance)) +
#       geom_boxplot() +
#       geom_jitter() +
#       facet_grid(Cohort~Taxon) +
#       scale_x_discrete(drop=FALSE) +
#       scale_y_log10() +
#       theme(axis.text.x = element_text(angle=90)) +
#       labs(x=element_blank(), y="Bacterial Load", title="Presence-Absence vs. Abundance")
#   }
#   
# map(list(prelimPresAbsAll, prelimPresAbsSubset), prelimPresAbsFigure)  
```

### Final prelim oral figure
```{r, echo=FALSE}
# prelimPresAbsAll <- DF %>%
#     left_join(totalAbundances, by="SampleID") %>%
#     gather("Taxon", "presAbs", c(G_vaginalis, C1, C2, C3, C4, C5, C6, A_vaginae, L_crispatus, L_gasseri, L_iners, L_jensenii, P_bivia, U_parvum, M_hominis)) %>%
#     filter(Taxon %in% c("G_vaginalis", "C1", "C6", "L_crispatus")) %>%
#     filter(Cohort=="Stanford") %>%
#     mutate(Taxon = factor(Taxon, levels=c("G_vaginalis", "C1", "C6", "L_crispatus"), labels=c("Gv", "C1", "C6", "Lc")),
#            presAbs=case_when(presAbs>0~"present", 
#                              presAbs==0~"absent"),
#            presAbs=factor(presAbs, levels=c("present", "absent")))
#   
#  prelimPresAbsSubset <-  prelimPresAbsAll %>%
#   filter(SampleID %in% subsetSamples)
# 
#   prelimPresAbsFigure <- function(df){
#       ggplot(data=df, aes(x=presAbs, y=totalAbundance)) +
#       geom_boxplot() +
#       geom_jitter() +
#       facet_grid(Cohort~Taxon) +
#       scale_x_discrete(drop=FALSE) +
#       scale_y_log10() +
#       theme(axis.text.x = element_text(angle=90)) +
#       labs(x=element_blank(), y="Bacterial Load", title="Presence-Absence vs. Abundance")
#   }
#   
# map(list(prelimPresAbsAll, prelimPresAbsSubset), prelimPresAbsFigure)  
```

F31 figure
```{r, echo=FALSE}
#function to calcualte spearman's rank correlation coefficients
# corTestFig1 <- function(df, name){
#   cor_test_results <- df %>%
#     filter(Cohort=="Stanford") %>%
#     split(list(.$Taxon1, .$Taxon2)) %>%
#       map(~cor.test(~absoluteAbundanceTax1+absoluteAbundanceTax2, data=.x, alternative="two.sided", method="spearman")) %>%
#       map(tidy) %>%
#       map2(., names(.), ~mutate(.x, Group=.y)) %>%
#       purrr::reduce(full_join, by = c("estimate", "statistic", "p.value", "method", "alternative", "Group")) %>%
#       separate(Group, c("Taxon1", "Taxon2"), sep="\\.") %>%
#       select(-p.value)
#   plot <- cor_test_results %>%
#       # mutate(p.value=case_when(p.value>0.05~"",
#       #                          p.value<=0.05&p.value>0.01~".",  
#       #                          p.value<=0.01&p.value>0.001~"*",
#       #                          p.value<=0.001&p.value>0.0004~"**",
#       #                          p.value<=0.0004&p.value>0.0001~"***",
#       #                          p.value<=0.0001~"****"),
#              Taxon1=factor(Taxon1, levels=c("C6", "C5", "C4", "C3", "C2", "C1", "Gv")), 
#              Taxon2=factor(Taxon2, levels=c("Lc", "Lg", "Lj", "Li"))) %>%
#     ggplot(aes(x=Taxon2, y=Taxon1, fill=estimate)) + #label=p.value
#       geom_tile() +
#       #geom_text() +
#       scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
#       scale_x_discrete(drop=FALSE) +
#       scale_y_discrete(drop=FALSE) +
#       labs(x="Lactobacillus", y="Gardnerella", title = "Abundance Correlations")
#   outputList <- list(Results=cor_test_results, Figure=plot)   
#     return(outputList)
# }
```

# Over time?
```{r, echo=FALSE}
# DF %>%
#   gather("taxon", "abundance", c(5:ncol(.))) %>%
#   #filter(taxon=="C1"|taxon=="L_crispatus") %>%
#   mutate(GDcoll=as.numeric(as.character(GDcoll))) %>%
# ggplot(aes(x=GDcoll, y=abundance, group=taxon, color=taxon)) +
#   geom_line() +
#   facet_wrap(~SubjectID) +
#   ylim(0,0.25) +
#   scale_x_continuous(breaks = waiver()) +
#   #scale_y_continuous(breaks = waiver()) +
#   theme(axis.text.x = element_text(angle=45))
```

```{r, echo=FALSE}
# DF %>%
#   gather("taxon", "abundance", c(5:ncol(.))) %>%
#   #filter(taxon=="C1"|taxon=="L_crispatus") %>%
#   mutate(GDcoll=as.numeric(as.character(GDcoll))) %>%
# ggplot(aes(x=GDcoll, y=abundance, group=taxon, color=taxon)) +
#   geom_line() +
#   facet_wrap(~Cohort) +
#   ylim(0,0.25) +
#   scale_x_continuous(breaks = waiver()) +
#   #scale_y_continuous(breaks = waiver()) +
#   theme(axis.text.x = element_text(angle=45))
```

# Session Info
```{r}
sessionInfo()
```