---
title: "Simulation of Gardnerella Gene Content Deconvolution"
author: "Hanna Berman"
date: "8/28/2019"
output: html_document
---

# Setup
## Packages
```{r}
library(tidyverse)
library(nnls)
library(glmnet)
set.seed(6789)
```

## File paths
```{r}
# #metadata
# sgDF <- file.path("..", "shotgunMetadata.tsv") %>%
#   read_tsv() %>%
#   mutate(SampleID=as.character(SampleID))
# 
# #metaphlan2 data
# load(file.path("/Users/hlberman/Documents/GitHub/pregnancy_metagenome/gard_mapping/clade_assignments/metaphlanAbundances.Rdata"))
# 
# #subset
# sampleSubset <- read_lines(file.path("..", "sampleSubset.txt"))

#humann2 outs
dataDir <- "/Volumes/GoogleDrive/My Drive/Callahan Lab/vagMicrobiomeMetagenome/humann2/out_tables"
GF_path <- file.path(dataDir, "VM_genefamilies_names_cpm.tsv")
# GFP_path <- file.path(dataDir, "VM_genefamilies_paths.tsv")
# GOnames_path <- file.path(dataDir, "map_go_name.txt")
# GFGO_path <- file.path(dataDir, "map_go_uniref90.txt")
# PWnames_path <- file.path(dataDir, "map_metacyc-pwy_name.txt") 
# PC_path <- file.path(dataDir, "VM_pathcoverage.tsv")
# PA_path <- file.path(dataDir, "VM_pathabundance.tsv")

# read in data frames
sgDF <- file.path("..", "shotgunMetadata.tsv") %>%
  read_tsv() %>%
  mutate(SampleID=as.character(SampleID))

#gardnerella reference sequences
roaryDir <- "/Volumes/GoogleDrive/My Drive/Callahan Lab/Gard WGS/20181114_reAnnotatedGvRoary70_set1"

geneAnnos <- file.path(roaryDir, "gene_presence_absence.csv") %>%
  read_csv() %>%
  select(Gene, Annotation)
print(geneAnnos)

GvRoary <- file.path(roaryDir, "gene_presence_absence.Rtab") %>%
  read_delim(delim="\t") %>%
  left_join(geneAnnos, by="Gene") %>%
  select(Gene, Annotation, everything())
print(GvRoary)

geneAnnos <- file.path(roaryDir, "gene_presence_absence.csv") %>%
  read_csv() %>%
  select(Gene, Annotation)
print(geneAnnos)

#Roary output for core genome
GvRoaryCore <- "/Volumes/GoogleDrive/My Drive/Callahan Lab/Gard WGS/20181008_GvRoary95strains/gene_presence_absence.csv" %>%
  read_csv()
print(GvRoaryCore)

#Reference dataframe
gvDF <- file.path("..", "GardnerellaMetadata.csv") %>%
  read_csv()
print(gvDF)
```

## Check Gardnerella read counts in samples
```{r}
GF <- read_delim(GF_path, delim = "\t") %>% # Gene family abundances
  separate(`# Gene Family`, into=c("geneFamily", "org"), sep="\\|", fill="right") %>%
  separate(org, into=c("Genus", "Species"), sep="\\.") %>%
  mutate(Genus=str_extract(Genus,"[^g__].*$"),
         Species=str_extract(Species, "[^s__].*"))

GF %>%
  dplyr::filter(Genus=="Gardnerella") %>%
  select(-geneFamily, -Genus, -Species) %>%
  colSums() %>%
  as.data.frame() %>%
  filter(.!=0) %>%
  summary(.)
```  
values for read counts: 10,000 (~1st q) 400,000 (~max)


# Basic Simulation with Gene Presence Absence
First simulate with gene presence absence information
## Create Samples
```{r}
sampleCreate <- function(genePA){
  x <- sample(c(1:6), 1)
  if(x==1){
    Sample <- genePA %>%
    dplyr::select(Gene, sample(c(3:ncol(.)), 1, replace=FALSE))}
  if(x==2){
    Sample <- genePA %>%
    select(Gene, sample(c(3:ncol(.)), 2, replace=FALSE))}
  if(x==3){
    Sample <- genePA %>%
    select(Gene, sample(c(3:ncol(.)), 3, replace=FALSE))}
  if(x==4){
    Sample <- genePA %>%
    select(Gene, sample(c(3:ncol(.)), 4, replace=FALSE))}
  if(x==5){
    Sample <- genePA %>%
    select(Gene, sample(c(3:ncol(.)), 5, replace=FALSE))}
  if(x==6){
    Sample <- genePA %>%
    select(Gene, sample(c(3:ncol(.)), 6, replace=FALSE))}
  return(Sample)
}

#subset reference strains
sampleStrains <- rerun(100, sampleCreate(GvRoary))
sampleStrains[1]

# #make sample abundances
# sampleAbundances <- c(1:100) %>%
#   map(~tibble(Strain=colnames(sampleStrains[[.x]])[2:ncol(sampleStrains[[.x]])],
#               Abundances=runif(ncol(sampleStrains[[.x]])-1))) %>% #abundances with normal dist. 
#   map(~mutate(.x, Abundances=Abundances/sum(Abundances)))
# sampleAbundances[1]
# 
# #sanity check that all sample abundances sum to 1
# sampleAbundances %>%
#   map(~sum(.x$Abundances)) %>%
#   unlist() %>%
#   map(~.x==1) %>%
#   unlist %>%
#   table
# 
# #read counts
# readCounts <- runif(100, min=10000, max=400000) %>%
#   round
# readCounts
#
# #Find new read counts per gene
# sampleStrainReads <- sampleAbundances %>%
#   map2(., readCounts, ~mutate(.x, strainTotalReads=Abundances*.y, 
#                                   sampleTotalReads=.y)) %>%
#   map(., ~left_join(.x, gvDF[,c("Strain", "Clade")], by="Strain")) %>%
#   map(., ~select(.x, Strain, Clade, Abundances, strainTotalReads, sampleTotalReads))
# sampleStrainReads[1]

#assign new read counts per gene
sampleGeneReads <- sampleStrains %>%
  map(., ~mutate_if(.x, is.numeric, ~.x*round(runif(1, 2000, 70000))))
sampleGeneReads[2]

# Get core genes
coreGenes <- GvRoaryCore %>%
  add_count(Annotation) %>%
  filter(`No. isolates`==72,
         n==1) %>%
  .$Annotation %>%
  {dplyr::filter(GvRoary, is.element(Annotation, .))} %>%
  .$Gene
coreGenes

#calculate relative abundance of each strain using marker genes
strainAb0 <- sampleGeneReads %>%
  map(~filter(.x, is.element(Gene, coreGenes))) %>%
  map(~summarise_if(.x, is.numeric, sum))
strainAb0[1]
samCoreTot0 <- sampleGeneReads %>%
  map(~filter(.x, is.element(Gene, coreGenes))) %>%
  map(~summarise_if(.x, is.numeric, sum)) %>%
  map(~summarise(.x, sums=rowSums(.x))) %>%
  map(~.x$sums) %>%
  unlist
samCoreTot0[1]

strainAbundances <- strainAb0 %>%
  map2(., samCoreTot0, ~mutate(.x, Totals=.y)) %>%
  map(., ~mutate_all(.x, ~.x/Totals)) %>%
  map(., ~select(.x, -Totals)) %>%
  map(., ~gather(.x, "Strain", "Abundance"))
strainAbundances[1]

#turn strain abundances into clade abundances and reduce to one table
cladeAbundances <- strainAbundances %>%
  map(., ~left_join(.x, gvDF[,c("Strain", "Clade")], by="Strain")) %>%
  map(., ~group_by(.x, Clade)) %>%
  map(., ~summarise(.x, Abundance=sum(Abundance))) %>%
  map(., ~ungroup(.x)) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>% # add sample labels
  map(., ~select(.x, Sample, Clade, Abundance)) %>%
  purrr::reduce(dplyr::full_join, by=c("Clade", "Abundance", "Sample")) %>% # reduce table
  spread(Clade, Abundance) %>%
  replace_na(list(`1`=0, `2`=0, `3`=0, `4`=0, `5`=0, `6`=0))
cladeAbundances
```

reference gene
```{r}
```


## Estimated coefficients
Using with alpha-L-fucosidase gene found only in clades 1 and 5
```{r}
#Create unlabeled gene reads for testing
testSamples <- sampleGeneReads %>%
  map(., ~gather(.x, "Strain", "Reads", 2:ncol(.x))) %>%
  map(., ~group_by(.x, Gene)) %>%
  map(., ~summarise(.x, Reads=sum(Reads)))
testSamples[2]

#get alpha-L-fucosidase gene name
aLfuc <- GvRoary %>%
  filter(str_detect(Annotation, "HMPREF0421_20101")) %>%
  .$Gene
aLfuc  

#G inverse (total reads/SecY reads):
#First get the total number of reads in each sample
samTot <- testSamples %>%
  map(~summarise(.x, samTots=sum(Reads))) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Sample", "samTots"))
samTot

#get the number of secY (universal single copy) reads  
Gs <- testSamples %>%
  map(., ~filter(.x, Gene=="secY")) %>%
  map(., ~rename(.x, secYreads=Reads)) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Sample", "secYreads")) %>%
  left_join(samTot, by="Sample") %>%
  mutate(Ginv=samTots/secYreads) %>%
  select(Sample, secYreads, samTots, Ginv)
Gs

#bind G column and normalize read counts
testGene <- testSamples %>%
  map(., ~filter(.x, Gene==aLfuc)) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Gene", "Reads", "Sample")) %>%
  left_join(cladeAbundances, by="Sample") %>%
  left_join(Gs[,c("Sample", "Ginv")], by="Sample") %>%
  mutate(GReads=Reads/Ginv)
testGene

glm(data=testGene, GReads~`1`+`2`+`3`+`4`+`5`+`6`) %>%
  summary()
```
The coefficients should be 1 and 0.


NNLS
make matrix A and b for nnls(A,b) to calculate $Ax=b$
```{r}
b <- testGene %>%
  .$GReads
A <- testGene %>%
  select(`1`,`2`,`3`,`4`,`5`,`6`) %>%
  as.matrix()
nnls(A, b)
```

LASSO
```{r}
test <- glmnet(A,b, family="gaussian", alpha = 1)
plot(test, xvar="lambda", label=TRUE)
coef(test,s=c(2.95, 2.96, 2.97, 2.98, 2.99, 3.05, 3.06))
test1 <- cv.glmnet(A, b, alpha=1)
test1$lambda.min
coef(test1, s="lambda.min" )
plot(test1)
```

### Redo G copy number/reference gene reads
Using with alpha-L-fucosidase gene found only in clades 1 and 5
```{r}
#Create unlabeled gene reads for testing
testSamples <- sampleGeneReads %>%
  map(., ~gather(.x, "Strain", "Reads", 2:ncol(.x))) %>%
  map(., ~group_by(.x, Gene)) %>%
  map(., ~summarise(.x, Reads=sum(Reads)))
testSamples[2]

#get alpha-L-fucosidase gene name
aLfuc <- GvRoary %>%
  filter(str_detect(Annotation, "HMPREF0421_20101")) %>%
  .$Gene
aLfuc  

#G inverse (total reads/SecY reads):
#First get the total number of reads in each sample
samTot <- testSamples %>%
  map(~summarise(.x, samTots=sum(Reads))) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Sample", "samTots"))
samTot

#get the number of secY (universal single copy) reads  
Gs <- testSamples %>%
  map(., ~filter(.x, Gene=="secY")) %>%
  map(., ~rename(.x, secYreads=Reads)) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Sample", "secYreads")) %>%
  left_join(samTot, by="Sample") %>%
  mutate(G=1/secYreads,
         Ginv=secYreads/1) %>%
  select(Sample, secYreads, samTots, G, Ginv)

#bind G column and normalize read counts
testGene <- testSamples %>%
  map(., ~filter(.x, Gene==aLfuc)) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Gene", "Reads", "Sample")) %>%
  left_join(cladeAbundances, by="Sample") %>%
  left_join(Gs[,c("Sample", "G", "Ginv")], by="Sample") %>%
  mutate(GReads=Reads*G)
testGene

glm(data=testGene, GReads~`1`+`2`+`3`+`4`+`5`+`6`) %>%
  summary()

glm(data=testGene, Reads~`1`*Ginv+`2`*Ginv+`3`*Ginv+`4`*Ginv+`5`*Ginv+`6`*Ginv) %>%
  summary()
```

```{r}
testSamples %>%
  map(., ~filter(.x, Gene==aLfuc)) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Gene", "Reads", "Sample")) %>%
  left_join(cladeAbundances, by="Sample") %>%
  left_join(Gs[,c("Sample", "G")], by="Sample") %>%
  mutate(GReads=Reads*G)
```


NNLS
make matrix A and b for nnls(A,b) to calculate $Ax=b$
```{r}
b <- testGene %>%
  .$GReads
A <- testGene %>%
  select(`1`,`2`,`3`,`4`,`5`,`6`) %>%
  as.matrix()
nnls(A, b)
```

LASSO
```{r}
test <- glmnet(A,b, family="gaussian", alpha = 1)
plot(test, xvar="lambda", label=TRUE)
test1 <- cv.glmnet(A, b, alpha=1)
test1$lambda.min
coef(test1, s="lambda.min" )
plot(test1)
```

## Actual coefficients
Calculated by actual read counts  
Total aLfucosidase reads = Clade1 aLf reads + Clade2 aLf reads + Clade3 aLf reads + Clade4 aLf reads + Clade5 aLf reads + Clade6 aLf reads
```{r}
geneCladeAbundance <- sampleGeneReads %>%
  map(., ~filter(.x, Gene==aLfuc)) %>%
  map(., ~gather(.x, "Strain", "Reads", 2:ncol(.x))) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Gene", "Strain", "Reads", "Sample")) %>%
  left_join(gvDF[,c("Strain", "Clade")], by="Strain") %>%
  group_by(Clade) %>%
  summarize(Reads=sum(Reads))
geneCladeAbundance

#reads of alpha L fucosidase per clade
x <- sampleGeneReads %>%
  map(., ~filter(.x, Gene==aLfuc)) %>%
  map(., ~gather(.x, "Strain", "Reads", 2:ncol(.x))) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Gene", "Strain", "Reads", "Sample")) %>%
  left_join(gvDF[,c("Strain", "Clade")], by="Strain") %>%
  group_by(Clade, Sample) %>%
  summarise(Reads=sum(Reads))
x

#reads of alpha L fucosidase per sample
y <- sampleGeneReads %>%
  map(., ~filter(.x, Gene==aLfuc)) %>%
  map(., ~gather(.x, "Strain", "Reads", 2:ncol(.x))) %>%
  map2(., c(1:100), ~mutate(.x, Sample=.y)) %>%
  purrr::reduce(full_join, by=c("Gene", "Strain", "Reads", "Sample")) %>%
  left_join(gvDF[,c("Strain", "Clade")], by="Strain") %>%
  group_by(Sample) %>%
  summarise(alfucReads=sum(Reads))
y

#join 
z <- x %>%
  left_join(y, by="Sample") %>%
  spread(Clade, Reads) %>%
  replace_na(list(`1`=0, `2`=0, `3`=0, `4`=0, `5`=0, `6`=0))
z

glm(data=z, alfucReads~`1`+`2`+`3`+`4`+`5`+`6`) %>%
  summary()
```

Session Info
```{r}
sessionInfo()
```

